 

### 2023-03-01 paper2.Rmd
Revisions to Paper 2 from KK, MM, SV, and JC

# Loading libraries
# library(MASS) # We load this and then detach it just to clear the package
# library(margins) # We load this and then detach it just to clear the package
# library(lme4) # We load this and then detach it just to clear the package
# detach("package:lme4", unload=TRUE) # Needed to use dplyr's select function
# detach("package:margins", unload=TRUE) # Needed to use dplyr's select function
# detach("package:MASS", unload=TRUE) # Needed to use dplyr's select function
library(readxl)
library(readr)
library(dplyr)
library(zoo)
library(tidyr)
library(psych)
library(ggplot2)
library(RColorBrewer)
library(sp)
library(broom)
library(sf)
library(scales)
library(viridis)
library(data.table)
library(ggpubr)
library(ggmap)
library(maps)
library(mapdata)
library(ggrepel)
library(table1)
library(stringr) 
library(stringi)
library(fixest)
library(stats)
library(lubridate)
library(forecast)
library(gtools)
# library(devtools)
library(reticulate)
library(stargazer)
library(lwgeom)
if (!require("ggsflabel")) devtools::install_github("yutannihilation/ggsflabel")
library(openxlsx)
library(gganimate)
library(transformr)
library(huxtable)
library(ggtext)
library(margins)
library(ncdf4) # package for netcdf manipulation
library(magrittr)
library(lmtest)
library(sandwich)
library(rgee)
library(gt)
library(knitr)
library(webshot2)
library(sysfonts)
library(showtext)
font_add_google("Source Sans Pro")
library(gridExtra)
library(tibble)
showtext_auto() 
showtext_opts(dpi = 300)  
# library(googledrive)
# drive_auth(email = TRUE)
library(rmdHelpers)
library(gtExtras)
library(kableExtra)
library(DescTools)
library(rvest)
library(httr)
library(polite)
library(janitor)
library(cowplot)
library(ggridges)
# library(ggsn)

@@

library(rayshader)

@@

@@

{\bf Background:} Climate change-related exposures (CCEs) have been increasingly linked to worsened mental health, yet few studies of CCEs have taken place in low- and middle-income countries.

{\bf Methods:} To better understand how climate conditions may be impacting population mental health in these settings, we leverage meteorologic, geospatial, and health system data reported by `r num_facilities_pretty` facilities in Madagascar from 2010 to 2020 to explore the mental health impacts of three exposures of climate using negative binomial regression: temperature, soil moisture, and duration of cyclones and tropical storms.

{\bf Findings and interpretation:} We found a statistically significant reduction in monthly cases associated with temperature—approximately `r function_dynamic_text_results(section = "main", coefficient_option = "exposure_temperature", model_option = 4, value = "irr", value_include = "only")` cases reported per each degree Celsius `r function_dynamic_text_results(section = "main", coefficient_option = "exposure_temperature", model_option = 4, value = "irr", value_include = TRUE)`. A multiverse analysis of `r num_models_multiverse` models found this estimate to be robust to alternative specifications, while a subgroup analysis by ecoregion indicated that this association was primarily localized to facilities in the central highlands, where cooler temperatures are more prevalent. We further found that higher soil moisture may be associated with higher case counts in the future, with an `r function_dynamic_text_results(section = "lags", lag_option = 3, coefficient_option = "exposure_sm", value = "irr", value_include = "only")`  elevation in incident cases reported `r function_dynamic_text_results(section = "lags", lag_option = 3, coefficient_option = "exposure_sm", value = "irr", value_include =TRUE)`  after a lag of three months. We further found that greater soil moisture was associated with higher cases reported in the eastern lowlands and the northern dry deciduous forests, which is consistent with the geographic concentration of flood risk in Madagascar.

@@

Among low-income countries most vulnerable to the health consequences of climate change is Madagascar, a nation that generates less than a tenth of a percentage of global carbon dioxide emissions. Worsening CCEs in Madagascar (primarily in the form of droughts, floods, and cyclones) are likely to exacerbate endemic, climate-sensitive threats to health, including malnutrition and malaria.[@rakotoarisonAssessmentRiskVulnerability2018] In particular, persistent drought conditions in recent years have led to extensive food insecurity.[@adamtaylorMadagascarHeadedClimate] Though there is debate concerning the extent to which human activity is directly responsible for Madagascar's current drought,[@harringtonAttributionSevereLow; @zhongManyDimensionsDrought2021] the extremity of the situation has led some observers to classify conditions in Madagascar as the first climate change famine in history.[@worldfoodprogrammeMadagascarSevereDrought2021]

@@

In this ecological study, we investigated the association of climate conditions and population mental health in Madagascar between 2010 and 2020 using routine health system data from Madagascar’s Ministry of Public Health. The use of health system data enabled us to indirectly evaluate the extent to which existing public health data may already be showing signals of CCEs, which may aid efforts to prevent and mitigate mental health stressors and better design systemic responses to mental health needs. As a measure of population mental health, we considered facility-level monthly counts of new diagnoses of mental disorders. To capture the importance of both variation in temperature and precipitation conditions in Madagascar, we explored the predictive association of three climate conditions with incident mental health disorders: mean monthly ambient air temperature, mean monthly soil moisture, and monthly cyclone activity.

@@

Exploratory data analysis indicated that the classification and reporting of counts of mental disorders in GESIS varied across types of facilities in their level of detail. Among CHD facilities, incidence of mental disorders was aggregated with neurological disorders (coded as “neuro psychiatric disorders”), while CHRR and CHU facilities reported more detailed counts of specific mental disorders, such as depression, personality disorders, and schizophrenia. Across all facilities, disaggregated data was reported by age group but not by sex. Among CSB facilities, incidences of mental disorders were reported by age group under “mental illnesses and psychic disorders” until June 2015, when reporting was reclassified to “mental disorders,” and disaggregated data was reported by both age and sex.

@@

To gauge the accuracy of the geolocating process, a subset of 1% of geolocated facilities was selected for validation against manually identified coordinates. Comparing the geolocated coordinates to validated coordinates for each facility, the mean distance between coordinates in the validation sample was `r format(round(as.numeric(validation_mean_dist),2), big.mark = ",")` km., with a maximum distance of approximately `r format(round(as.numeric(validation_max_dist), 1), big.mark = ",")` kilometers, indicating high accuracy in the geolocated process relative to the spatial resolution of climate data. Additional detail on the geolocating and validation process is available in Appendix 2.2.

@@

ERA 5 is a global atmospheric reanalysis from the European Centre for Medium-Range Weather Forecasts (ECMWF) and is freely available through its Copernicus Climate Change Service (C3S)@europeanspaceagencyEuropeanSpaceAgency2022. The spatial resolution of ERA 5 is approximately 30 km., with quality-assured hourly estimates available from 1959 to the present  @europeanspaceagencyEuropeanSpaceAgency2022. Using Google Earth Engine, a time series of monthly aggregate values for mean air temperature in degrees Kelvin at a height of 2 meters was obtained for each facility geolocation  @ERA5MonthlyAggregates2022. Temperature data was then converted to degrees Celsius.

We obtained soil moisture data from the European Space Agency’s soil moisture dataset (ESA CCI SM), the world’s first and most comprehensive, multi-decade, satellite-observed dataset of global soil moisture. The spatial resolution of ESA CCI SM is approximately 26 km., with daily estimates available from 1978 to 2021. Using ESA CCI SM active and passive radiometer data, a time series of monthly aggregate values for mean soil moisture—measured as saturation percentage—was constructed from daily observations at each land grid point within Madagascar. We then applied nearest neighbor matching to link these time series to our geolocated facilities.

@@

To test for lagged associations—that is, whether in a given month, climate conditions in prior months were associated with counts of mental disorders—versions of Model 4 were conducted using lagged predictors. Measures of association were estimated for each monthly lag increment between 1 month and 18 months, inclusive.

@@

First, we considered the role of data cleaning in the outcome data and chose to compare our analysis using the cleaned outcome data against the raw, unadjusted values. Second, we chose to compare our findings using data from all facilities (the full sample) against a sample without district and regional hospital (CSB/FSP facilities only). Primary care facilities have smaller catchment areas than hospitals; consequently, it is more plausible that climate data for primary care facilities will closely reflect climate exposure to patients compared to referral hospitals. Third, we considered the decision to model temperature and soil moisture as an additive or interactive relationship (Model 3 vs. Model 4), with both alternatives included in our multiverse analysis. Fourth, we evaluated our decision to account for overdispersion with a NB model by using a Poisson regression framework as an alternative. Fifth, we evaluated alternative specifications of exposure to cyclone activity, using three options: our binary indicator, our count indicator, and no term for cyclone activity included. Sixth, we toggled whether time fixed effects (month and year) were included. Seventh, we similarly toggled regional fixed effects.

@@

model_outputs_preferred_model %>%
  filter(model_type == "NB") %>%
  filter(coefficient %in% c("exposure_temperature:exposure_sm", "exposure_temperature", "exposure_sm", "days_with_cyclone")) %>%
  mutate(
    cell_coef =
      paste0(
        signif(exp(estimate), 3),
        " (",
        signif(exp(lower_99), 3),
        " – ",
        signif(exp(upper_99), 3),
        ")",
        function_pval_stars(p_value)
      ),
    cell_me = ifelse(
      coefficient == "exposure_temperature:exposure_sm",
      "",
      paste0(
        signif(ame, 3),
        " (",
        signif(ame_lower_99, 3),
        " – ",
        signif(ame_upper_99, 3),
        ")",
        function_pval_stars(ame_p_value)
      )
    ),
    coefficient = ifelse(
      coefficient == "exposure_sm",
      "Soil moisture",
      ifelse(
        coefficient == "exposure_temperature",
        "Temperature",
        ifelse(coefficient == "days_with_cyclone",
               "Cyclone (count)",
               "Temperature X Soil moisture"
        )
      )
    ),
    coefficient = factor(
      coefficient ,
      # Reordering group factor levels
      levels = c("Temperature",
                 "Soil moisture",
                 "Temperature X Soil moisture",
                 "Cyclone (count)")
    ),
    model = ifelse(
      model_form == "Temperature only",
      "Model 1 ",
      ifelse(
        model_form == "Soil moisture only",
        "Model 2 ",
        ifelse(model_form == "Temperature + Soil moisture",
               "Model 3 ",
               "Model 4 ")
      )
    ),
    model = factor(model ,      # Reordering group factor levels
                   levels = c("Model 1 ",
                              "Model 2 ",
                              "Model 3 ",
                              "Model 4 "))
  ) %>%
  dplyr::select(model, coefficient, cell_coef, cell_me) %>%
  rename('IRR (99% CI)' = cell_coef,
         'M.E. (99% CI)' = cell_me) %>%
  gt(rowname_col = "coefficient", groupname_col = "model") %>%
  tab_header(title = md("Table 2.2: incidence rate ratios (IRRs) and average marginal effects (AME) from Models 1 through 4.")) %>%

  cols_align(align = "left",
             columns = c('IRR (99% CI)' , 'M.E. (99% CI)')) %>%
  tab_footnote(footnote = "All models include month, year, and regional fixed effects.",
               locations = cells_row_groups()) %>%
  tab_footnote(footnote = "*** p < 0.001, ** p < 0.01, * p < 0.05",
               locations = cells_row_groups()) %>%
  opt_footnote_marks("letters") %>%
  tab_options(table.font.names = "Source Sans Pro") %>%
  tab_style(locations = cells_column_labels(),
            style     = list(cell_text(weight = "bold"))) %>%
  tab_options(table.width = pct(100))

@@

With respect to soil moisture, while our primary analysis found no evidence of a statistically significant association with reported cases in the full sample, we did find evidence of heterogeneity in our subgroup analysis. Among facilities in the highland subhumid forests, we found that each percentage point increase in soil moisture was associated with `r function_dynamic_text_results(section = "regional", coefficient_option = "exposure_sm", region_option = "2", value = "irr", value_include = "only")` reported cases `r function_dynamic_text_results(section = "regional", coefficient_option = "exposure_sm", region_option = "2", value = "irr", value_include = TRUE)`, with an average marginal effect of `r function_dynamic_text_results(section = "regional", coefficient_option = "exposure_sm", region_option = "2", value = "ame", value_include = "only")` reported cases `r function_dynamic_text_results(section = "regional", coefficient_option = "exposure_sm", region_option = "2", value = "ame", value_include = FALSE)`. Conversely, each percentage point increase in soil moisture in the dry deciduous forests was associated with a `r function_dynamic_text_results(section = "regional", coefficient_option = "exposure_sm", region_option = "4", value = "irr", value_include = "only")` increase in reported cases `r function_dynamic_text_results(section = "regional", coefficient_option = "exposure_sm", region_option = "4", value = "irr", value_include = TRUE)`, with an average marginal effect of `r function_dynamic_text_results(section = "regional", coefficient_option = "exposure_sm", region_option = "4", value = "ame", value_include = "only")` reported cases `r function_dynamic_text_results(section = "regional", coefficient_option = "exposure_sm", region_option = "4", value = "ame", value_include = FALSE)`.

@@

In our study of health facilities in Madagascar, we investigated how temperature, soil moisture, and exposure to cyclone activity impacted monthly incident cases of mental and neurological disorders reported by `r num_facilities_pretty`  facilities between 2010 and 2020. We found evidence of a statistically significant reduction in reported cases associated with mean monthly temperature, with warmer temperature being, on average, associated with 2 to 3 fewer cases per month. According to our multiverse analysis, the statistically significant inverse association we observed was robust to numerous specifications, while our subgroup analysis indicated that this association was primarily localized in the highland subhumid forest ecoregion.

With respect to soil moisture, we found evidence to suggest potential lagged impacts, with higher soil saturation being associated with an `r function_dynamic_text_results(section = "lags", lag_option = 3, coefficient_option = "exposure_sm", value = "irr", value_include = "only")` elevation in reported case counts in the subsequent quarter. We also found evidence of possible effect heterogeneity, with each percentage point increase in soil moisture leading to an *elevation* of cases in the dry deciduous forest ecoregion `r function_dynamic_text_results(section = "regional", coefficient_option = "exposure_sm", region_option = "4", value = "irr", value_include = TRUE)` and a *decrease* in cases in the highland subhumid forests `r function_dynamic_text_results(section = "regional", coefficient_option = "exposure_sm", region_option = "2", value = "irr", value_include = TRUE)`.

Finally, while our lagged models and multiverse analysis suggested the possibility of a negative relationship between cyclone activity and incident case counts reported in Madagascar, the marginal effects associated with cyclone duration were negligible, and we did not observe sufficient statistically significant evidence to reject the null assumption of no association at a level of 0.01.

It is important to note in our interpretation of these findings that the true relationship between reported mental health cases and actual cases of mental disorders is not well understood, particularly in Madagascar where data on disease etiology is limited. The changes in reported cases could reflect changes in the actual cases of mental disorders, but could also result from changes in help-seeking behavior, diagnostic capacity, or other factors. Additionally, it is widely recognized that a significant proportion of individuals with mental disorders do not receive any treatment, particularly in low-income settings.

@@

Though incidence rate ratios and marginal effect estimates associated with soil moisture were not statistically significant in our main analysis, we did observe a significant association of higher soil moisture and cases of mental health complaints in the following quarter, echoing similar findings from U.S. meteorological data linking higher precipitation with worsened mental health.[@obradovichEmpiricalEvidenceMental2018] In addition, our subgroup analysis also indicated significant and diverging effects among different ecoregions. In particular, we found that greater soil moisture was associated with higher cases reported in the eastern lowlands and the northern dry deciduous forests, which echoes the geographic concentration of flood risk in Madagascar.[@rakotoarisonSpatialAnalysisRisks2021] Conversely, the observed (though statistically modest) reduction in cases associated with higher soil moisture in the desert and xeric shrublands overlaps with the concentration of drought risk, where increased soil moisture could be taken as an indicator of reduced drought vulnerability. It is important to note, however, that the time frame of available GESIS data excludes the most severe period of the ongoing drought in Madagascar, limiting our ability to directly study the impact of the current crisis on population mental health in this analysis. Further research could leverage spatial reanalysis data on precipitation to further characterize these associations.

@@

Despite these limitations, our analysis brings together health data from over three thousand health facilities over ten years, utilizing the most comprehensive sources of climate data available, and linking the two via a rigorous, multistep geolocating process. This study—one of the only to investigate mental health and climate in Africa using national HMIS data—underscores the need for further research to understand the relationships and pathways between changing climate and population mental health.

>>>>

# Loading libraries
# library(MASS) # We load this and then detach it just to clear the package
# library(margins) # We load this and then detach it just to clear the package
# library(lme4) # We load this and then detach it just to clear the package
# detach("package:lme4", unload=TRUE) # Needed to use dplyr's select function
# detach("package:margins", unload=TRUE) # Needed to use dplyr's select function
# detach("package:MASS", unload=TRUE) # Needed to use dplyr's select function
library(readxl)
library(readr)
library(dplyr)
library(zoo)
library(tidyr)
library(psych)
library(ggplot2)
library(RColorBrewer)
library(sp)
library(broom)
library(sf)
library(scales)
library(viridis)
library(data.table)
library(ggpubr)
library(ggmap)
library(maps)
library(mapdata)
library(ggrepel)
library(table1)
library(stringr) 
library(stringi)
library(fixest)
library(stats)
library(lubridate)
library(forecast)
library(gtools)
# library(devtools)
library(reticulate)
library(stargazer)
library(lwgeom)
if (!require("ggsflabel")) devtools::install_github("yutannihilation/ggsflabel")
library(openxlsx)
library(gganimate)
library(transformr)
library(huxtable)
library(ggtext)
library(margins)
library(ncdf4) # package for netcdf manipulation
library(magrittr)
library(lmtest)
library(sandwich)
library(rgee)
library(gt)
library(knitr)
library(webshot2)
library(sysfonts)
library(showtext)
font_add_google("Source Sans Pro")
library(gridExtra)
library(tibble)
showtext_auto() 
showtext_opts(dpi = 300)  
# library(googledrive)
# drive_auth(email = TRUE)
library(rmdHelpers)
library(gtExtras)
library(kableExtra)
library(DescTools)
library(rvest)
library(httr)
library(polite)
library(janitor)
library(cowplot)
library(ggridges)
library(elevatr)
library(raster)
library(rayshader)

# library(ggsn)

@@

@@

ecological_zones <- read_excel(
  paste(data_climate_path,
        "ecological_zones.xlsx",
        sep = "/")) %>%
  rename(ecological_zone = 'ecological_zone (1 = Eastern rainforest; 2 = High Plateau; 3 = Arid/Desert; 4 = Dry Deciduos Forest)',
         REGION_NAM = region_name)

@@

{\bf Background:} Climate change-related exposures (CCEs) are linked to worsened mental health, yet few studies of CCEs have taken place in low- and middle-income countries.

{\bf Methods:} To better understand climate conditions and population mental health in these settings, we leverage meteorologic, geospatial, and monthly cases of mental and neurological disorders reported by `r num_facilities_pretty` facilities in Madagascar from 2010 to 2020 to explore the mental health impacts of three CCEs using negative binomial regression: temperature, soil moisture, and duration of cyclones and tropical storms.

{\bf Findings and interpretation:} We found a statistically significant reduction in monthly cases reported associated with temperature—approximately `r function_dynamic_text_results(section = "main", coefficient_option = "exposure_temperature", model_option = 4, value = "irr", value_include = "only")` cases reported per each degree Celsius increase `r function_dynamic_text_results(section = "main", coefficient_option = "exposure_temperature", model_option = 4, value = "irr", value_include = TRUE)`. A multiverse analysis of `r num_models_multiverse` models found this estimate to be robust to alternative specifications; ecoregion analysis indicated that this association was primarily localized to facilities in the central highlands, where cooler temperatures are more prevalent. We further found that higher soil moisture may be associated with higher case counts in the future, with an `r function_dynamic_text_results(section = "lags", lag_option = 3, coefficient_option = "exposure_sm", value = "irr", value_include = "only")`  elevation in incident cases reported `r function_dynamic_text_results(section = "lags", lag_option = 3, coefficient_option = "exposure_sm", value = "irr", value_include =TRUE)`  after a three month lag. Ecoloregion analysis suggest higher cases are reported in the eastern lowlands and the northern dry deciduous forests when soil moisture rises, which is consistent with the geographic concentration of flood risk in Madagascar. Our findings highlight the complex interactions between CCEs and population mental health and suggest potential heterogeneity in how changing climate conditions will impact mental health needs and system responses.

@@

Among low-income countries most vulnerable to the health consequences of climate change is Madagascar, a nation that generates one percent *of* one percent of global carbon dioxide emissions. Worsening CCEs in Madagascar (primarily in the form of droughts, floods, and cyclones) are likely to exacerbate endemic, climate-sensitive threats to health, including malnutrition and malaria.[@rakotoarisonAssessmentRiskVulnerability2018] In particular, persistent drought conditions in recent years have led to extensive food insecurity.[@adamtaylorMadagascarHeadedClimate] Though there is debate concerning the extent to which human activity is directly responsible for Madagascar's current drought,[@harringtonAttributionSevereLow; @zhongManyDimensionsDrought2021] the extremity of the situation has led some observers to classify conditions in Madagascar as the first climate change famine in history.[@worldfoodprogrammeMadagascarSevereDrought2021]

@@

In investigating the impact of climate exposures on mental health, health system data can serve as a useful tool in settings where epidemiological survey data is sparse. Despite the poor availability of mental health services in many settings, routine health system data can still provide valuable insights into trends and patterns of service use and the need and distribution of care for mental health issues. Although these data may only capture a share of the true burden of disease, reported cases and visits related can be a valuable starting point for preliminary research into the impact of climate change on population mental health in resource-constrained settings.

In this ecological study, we investigated the association of climate conditions and population mental health in Madagascar between 2010 and 2020 using routine health system data from Madagascar’s Ministry of Public Health. The use of health system data enabled us to indirectly evaluate the extent to which existing public health data may already be showing signals of CCEs, which may aid efforts to prevent and mitigate mental health stressors and better design systemic responses to mental health needs.

As a measure of population mental health, we considered facility-level monthly reports cases of mental disorders; in terms of exposure, we explored the predictive association of three CCEs with incident mental health visits: mean monthly ambient air temperature, mean monthly soil moisture, and monthly cyclone activity. As exposure to these stressors could involve a delay in onset of mental health symptoms and subsequent care seeking, we further explored whether exposures in prior months were associated with monthly reported cases in sequent ones.  We further considered interactive and additive models to account for complexity in the interaction of climate variables in Madagascar, as well as ecological variation to capture differences in regional climate systems.

@@

Exploratory data analysis indicated that the classification and reporting of counts of mental disorders in GESIS varied across types of facilities in their level of detail. Among CHD facilities, incidence of mental disorders was aggregated with neurological disorders (coded as “neuro psychiatric disorders”), while CHRR and CHU facilities reported more detailed counts of specific mental disorders, such as depression, personality disorders, and schizophrenia. Across all facilities, disaggregated data was reported by age group but not by sex. Among CSB facilities, incidences of mental disorders were reported by age group under “mental illnesses and psychic disorders” until June 2015, when reporting was reclassified to “mental disorders,” and disaggregated data was reported by both age and sex. A summary of classification by facility type and year is provided in Appendix 2.1, Table S2.1, with English translations of the French disease names provided in Table S2.2. and yearly cumulative reported case totals by disorder reported by facility level and year in Tables S2.3-5.

@@

To gauge the accuracy of the geolocating process, a subset of 1% of geolocated facilities was selected for validation against manually identified coordinates. Comparing the geolocated coordinates to validated coordinates for each facility, the mean distance between coordinates in the validation sample was `r format(round(as.numeric(validation_mean_dist),2), big.mark = ",")` km, with a maximum distance of approximately `r format(round(as.numeric(validation_max_dist), 1), big.mark = ",")` km, indicating high accuracy in the geolocated process relative to the spatial resolution of climate data. Additional detail on the geolocating and validation process is available in Appendix 2.2.

@@

ERA 5 is a global atmospheric reanalysis from the European Centre for Medium-Range Weather Forecasts (ECMWF) and is freely available through its Copernicus Climate Change Service (C3S)@europeanspaceagencyEuropeanSpaceAgency2022. The spatial resolution of ERA 5 is approximately 30 km, with quality-assured hourly estimates available from 1959 to the present  @europeanspaceagencyEuropeanSpaceAgency2022. Using Google Earth Engine, a time series of monthly aggregate values for mean air temperature in degrees Kelvin at a height of 2 meters was obtained for each facility geolocation  @ERA5MonthlyAggregates2022. Temperature data was then converted to degrees Celsius.

We obtained soil moisture data from the European Space Agency’s soil moisture dataset (ESA CCI SM), the world’s first and most comprehensive, multi-decade, satellite-observed dataset of global soil moisture. The spatial resolution of ESA CCI SM is approximately 26 km, with daily estimates available from 1978 to 2021. Using ESA CCI SM active and passive radiometer data, a time series of monthly aggregate values for mean soil moisture—measured as saturation percentage—was constructed from daily observations at each land grid point within Madagascar. We then applied nearest neighbor matching to link these time series to our geolocated facilities.

@@

Though CCEs could acutely impact mental health in our sample, we could anticipate that these impacts might not immediately result in changes to monthly case and visit counts, owing to delays in symptom onset, care seeking, and care delivery. To test for lagged associations—that is, whether in a given month, climate conditions in prior months were associated with counts of mental disorders—versions of Model 4 were conducted using lagged predictors. Measures of association were estimated for each monthly lag increment between 1 month and 18 months, inclusive.

@@

First, we considered the role of data cleaning in the outcome data and chose to compare our analysis using the cleaned outcome data against the raw, unadjusted values. Second, we chose to compare our findings using data from all facilities (the full sample) against a sample without district and regional hospital (CSB/FSP facilities only). Primary care facilities have smaller catchment areas than hospitals; consequently, it is more plausible that climate data for primary care facilities will closely reflect climate exposure to patients compared to referral hospitals. Third, we considered the decision to model temperature and soil moisture as an additive or interactive relationship (Model 3 vs. Model 4), with both alternatives included in our multiverse analysis. Fourth, we evaluated our decision to account for overdispersion with a NB model by using a Poisson regression framework as an alternative. Fifth, we evaluated alternative specifications of exposure to cyclone activity, using three options: our binary indicator, our count indicator, and no term for cyclone activity included. Sixth, we varied whether time fixed effects (month and year) were included. Seventh, we similarly included and excluded regional fixed effects for comparison.

@@

model_outputs_preferred_model %>%
  filter(model_type == "NB") %>%
  filter(coefficient %in% c("exposure_temperature:exposure_sm", "exposure_temperature", "exposure_sm", "days_with_cyclone")) %>%
  mutate(
    cell_coef =
      paste0(
        signif(exp(estimate), 3),
        " (",
        signif(exp(lower_99), 3),
        " – ",
        signif(exp(upper_99), 3),
        ")",
        function_pval_stars(p_value)
      ),
    cell_me = ifelse(
      coefficient == "exposure_temperature:exposure_sm",
      "",
      paste0(
        signif(ame, 3),
        " (",
        signif(ame_lower_99, 3),
        " – ",
        signif(ame_upper_99, 3),
        ")",
        function_pval_stars(ame_p_value)
      )
    ),
    coefficient = ifelse(
      coefficient == "exposure_sm",
      "Soil moisture",
      ifelse(
        coefficient == "exposure_temperature",
        "Temperature",
        ifelse(coefficient == "days_with_cyclone",
               "Cyclone (count)",
               "Temperature X Soil moisture"
        )
      )
    ),
    coefficient = factor(
      coefficient ,
      # Reordering group factor levels
      levels = c("Temperature",
                 "Soil moisture",
                 "Temperature X Soil moisture",
                 "Cyclone (count)")
    ),
    model = ifelse(
      model_form == "Temperature only",
      "Model 1 ",
      ifelse(
        model_form == "Soil moisture only",
        "Model 2 ",
        ifelse(model_form == "Temperature + Soil moisture",
               "Model 3 ",
               "Model 4 ")
      )
    ),
    model = factor(model ,      # Reordering group factor levels
                   levels = c("Model 1 ",
                              "Model 2 ",
                              "Model 3 ",
                              "Model 4 "))
  ) %>%
  dplyr::select(model, coefficient, cell_coef, cell_me) %>%
  rename('IRR (99% CI)' = cell_coef,
         'M.E. (99% CI)' = cell_me) %>%
  gt(rowname_col = "coefficient", groupname_col = "model") %>%
  tab_header(title = md("Table 2.2: incidence rate ratios (IRRs) and average marginal effects (AME) of climate exposures on monthly reported cases of mental and neurological disorders, Models 1 through 4.")) %>%

  cols_align(align = "left",
             columns = c('IRR (99% CI)' , 'M.E. (99% CI)')) %>%
  tab_footnote(footnote = "Model 1: Temperature only. Model 2: Soil moisture only. Model 3: Temperature and soil moisture, additive model. Model 4: Temperature and soil moisture, interactive model. All models include a count indicator for days of exposure to cyclone activity, as well as month, year, and regional fixed effects.",
               locations = cells_row_groups()) %>%
  tab_footnote(footnote = "*** p < 0.001, ** p < 0.01, * p < 0.05",
               locations = cells_row_groups()) %>%
  opt_footnote_marks("letters") %>%
  tab_options(table.font.names = "Source Sans Pro") %>%
  tab_style(locations = cells_column_labels(),
            style     = list(cell_text(weight = "bold"))) %>%
  tab_options(table.width = pct(100))

@@

With respect to soil moisture, while our primary analysis found no evidence of a statistically significant association with reported cases in the full sample, we found evidence of heterogeneity in our subgroup analysis. Among facilities in the highland subhumid forests, we found that each percentage point increase in soil moisture was associated with `r function_dynamic_text_results(section = "regional", coefficient_option = "exposure_sm", region_option = "2", value = "irr", value_include = "only")` reported cases `r function_dynamic_text_results(section = "regional", coefficient_option = "exposure_sm", region_option = "2", value = "irr", value_include = TRUE)`, with an average marginal effect of `r function_dynamic_text_results(section = "regional", coefficient_option = "exposure_sm", region_option = "2", value = "ame", value_include = "only")` reported cases `r function_dynamic_text_results(section = "regional", coefficient_option = "exposure_sm", region_option = "2", value = "ame", value_include = FALSE)`. Conversely, each percentage point increase in soil moisture in the dry deciduous forests was associated with a `r function_dynamic_text_results(section = "regional", coefficient_option = "exposure_sm", region_option = "4", value = "irr", value_include = "only")` increase in reported cases `r function_dynamic_text_results(section = "regional", coefficient_option = "exposure_sm", region_option = "4", value = "irr", value_include = TRUE)`, with an average marginal effect of `r function_dynamic_text_results(section = "regional", coefficient_option = "exposure_sm", region_option = "4", value = "ame", value_include = "only")` reported cases `r function_dynamic_text_results(section = "regional", coefficient_option = "exposure_sm", region_option = "4", value = "ame", value_include = FALSE)`.

@@

In our study of health facilities in Madagascar, we investigated how temperature, soil moisture, and exposure to cyclone activity impacted monthly incident cases of mental and neurological disorders reported by `r num_facilities_pretty`  facilities between 2010 and 2020. This study had three major findings. First, we found evidence of a statistically significant reduction in reported cases of approximately 4.5% associated with mean monthly temperature, with warmer temperature being, on average, associated with 2 to 3 fewer cases per month. According to our multiverse analysis, the statistically significant inverse association we observed was robust to numerous specifications, while our subgroup analysis indicated that this association was primarily localized in the highland subhumid forest ecoregion. Second, we found evidence to suggest potential lagged impacts, with higher soil saturation being associated with an `r function_dynamic_text_results(section = "lags", lag_option = 3, coefficient_option = "exposure_sm", value = "irr", value_include = "only")` elevation in reported case counts in the subsequent quarter. We also found evidence of possible effect heterogeneity, with each percentage point increase in soil moisture leading to an *elevation* of cases in the dry deciduous forest ecoregion and a *decrease* in cases in the highland subhumid forests. Third, while our lagged models and multiverse analysis suggested the possibility of a negative relationship between cyclone activity and incident case counts reported in Madagascar, the marginal effects associated with cyclone duration were negligible, and we did not observe sufficient statistically significant evidence to reject the null assumption of no association at a level of 0.01.

It is important to note in our interpretation of these findings that the true relationship between reported mental health cases and actual cases of mental disorders is not well understood, particularly in Madagascar, where there is no epidemiological data on the prevalence or distribution of mental disorders in the general population. The changes in reported cases could reflect changes in the actual cases of mental disorders, but could also result from the influence of CCEs on help-seeking behavior, diagnostic capacity, or other factors. Additionally, it has been widely recognized that the majority individuals with mental disorders often do not receive any mental health treatment, particularly in low-income settings. [@demyttenaerePrevalenceSeverityUnmet2004]

@@

Though incidence rate ratios and marginal effect estimates associated with soil moisture were not statistically significant in our main analysis, we observed a significant association of higher soil moisture and cases of mental health complaints in the following quarter, echoing similar findings from U.S. meteorological data linking higher precipitation with worsened mental health.[@obradovichEmpiricalEvidenceMental2018] In addition, our subgroup analysis also indicated significant and diverging effects among different ecoregions. In particular, we found that greater soil moisture was associated with higher cases reported in the eastern lowlands and the northern dry deciduous forests, which echoes the geographic concentration of flood risk in Madagascar.[@rakotoarisonSpatialAnalysisRisks2021] Conversely, the observed (though statistically modest) reduction in cases associated with higher soil moisture in the desert and xeric shrublands overlaps with the concentration of drought risk, where increased soil moisture could be taken as an indicator of reduced drought vulnerability. It is important to note, however, that the time frame of available GESIS data excludes the most severe period of the ongoing drought in Madagascar, limiting our ability to directly study the impact of the current crisis on population mental health in this analysis. Further research could leverage spatial reanalysis data on precipitation to further characterize these associations.

@@

Despite these limitations, our study has important implications for future research on the relationship between climate change exposures and population mental health. By leveraging, reported cases from over three thousand health facilities over ten years, utilizing the most comprehensive sources of climate data available, and linking the two via a rigorous, multistep geolocating process, our analysis demonstrates how routine health system data may offer valuable insights, particularly when other epidemiological data are not available. As one of the only studies to investigate mental health and climate in Africa using national HMIS data, our findings also underscore the need for further research to understand the relationships and pathways between changing climate and population mental health, particularly countries most vulnerable to climate change. Finally, our findings and interpretations highlight the importance of local knowledge and collaboration with local ministries in global health research, to make best use of existing data and to foster research to support health system strengthening. We believe that this approach to research is essential for understanding the complex relationships and mechanisms linking climate change exposures to mental health, as well as identifying strategies to mitigate these impacts amid the unfolding global climate crisis.
